Binary companions of evolved stars in \apogee\ \DR: \\ Search method and catalog of $\sim$5,000 companions

### 2018-04-06 paper/1-catalog/twoface-catalog.tex
Affil
M not K

% to-do list
% ----------
% - Make sure we don't use $K$ as a number of samples - that's $M$ - because $K$
%   is the velocity semi-amplitude!
% - Figure out what to say about upper RGB stars / coherent oscillations!

@@

% other notes
% -----------
% - Binary engulfment reference: https://arxiv.org/pdf/1002.2216.pdf

@@

% APOGEE RV var co-authors
\author{Nathan De Lee}
\affiliation{Department of Physics and Astronomy, Vanderbilt University, VU Station 1807, Nashville, TN 37235, USA}

@@

\author{Nicholas Troup}
\affiliation{}

@@

\begin{abstract}\noindent % trust me
% Context
Multi-epoch radial velocity measurements of stars can be used to identify
stellar, sub-stellar, and planetary-mass companions.
Even a small number of observation epochs can be informative about
companions, though there can be multiple qualitatively different orbital
solutions that fit the data.
We have custom-built a Monte Carlo sampler (\thejoker) that delivers reliable
(and often highly multi-modal) posterior samplings for companion orbital
parameters given sparse radial-velocity data.
% Aims
Here we use \thejoker\ to perform a search for companions to \nstars\ red-giant
stars observed in the \apogee\ survey (\DR) with $\geq 3$ spectroscopic epochs.
% Methods
We select stars with probable companions by making a cut on our posterior belief
about the amplitude of the stellar radial-velocity variation induced by the
orbit.
% Results
We provide (1)~a catalog of \nunimodal\ companions for which the stellar
companion properties can be confidently determined, (2)~a catalog of \nhighK\
stars that likely have companions, but would require more observations to
uniquely determine the orbital properties, and (3)~posterior samplings for the
full orbital parameters for all stars in the parent sample.
We highlight interesting systems and show the characteristics of systems with
confidently determined companion properties.
\end{abstract}

@@

Related to the above, we assume that all RV variations are
gravitational---\emph{Kepler}.
We see tentative evidence for surface oscillations at the upper giant branch
(see \figurename~\ref{fig:P-Teff-logg-e}, lower left), which we presently treat
as excess variance or intrinsic jitter.
Further observations are needed to determine whether these ``systems'' have
incorrect stellar parameters, or indeed show surface oscillations related to
asteroseismic modes.

@@

\software{
    \package{Astropy} (\citealt{Astropy-Collaboration:2013}),
    \package{emcee} (\citealt{Foreman-Mackey:2013}),
    \package{IPython} (\citealt{Perez:2007}),
    \package{matplotlib} (\citealt{Hunter:2007}),
    \package{numpy} (\citealt{Van-der-Walt:2011}),
    \package{scikit-learn} (\citealt{Pedregosa:2011}),
    \package{scipy} (\url{https://www.scipy.org/}),
    \package{schwimmbad} (\citealt{Price-Whelan:2017a}),
    \package{sqlalchemy} (\url{https://www.sqlalchemy.org/}),
    \package{thejoker} (\citealt{Price-Whelan:2017b}),
    \package{twobody} (\todo{twobody zenodo}).
}

@@

For each $n$ of $N$ RGB stars in APOGEE, we obtain $K$ posterior samples over
primary orbital parameters $\bs{\theta} = (P, e, \omega, M_0, K, v_0)$ and the
excess-variance parameter, $y = \ln s^2$, using \thejoker; For brevity in
expressions below, we will use the vector
\begin{equation}
    \bs{w} = (\bs{\theta}, y)
\end{equation}
to represent the full set of parameters.
To obtain this sampling, we use an interim (Gaussian) prior on the
excess-variance parameter parametrized by a mean and standard deviation, i.e.
$\alpha_0 = (\mu_{y,0}, \sigma_{y,0})$ as described above.
For a given source, the posterior samples in the above parameters are drawn from
the distribution
\begin{equation}
    \bs{w}_k \sim p(\bs{w}_k \given D, \bs{\alpha}_0)
\end{equation}
where $D$ represents the data for a given object.

We want to compute the likelihood of all data from all $N$ stars, $\{D_n\}$,
given a new set of hyperparameters $\bs{\alpha}$
\begin{equation}
    p(\{D_n\} \given \bs{\alpha}) = \prod_n^N p(D_n \given \bs{\alpha})
\end{equation}
where in the above, we have assumed that this likelihood is separable ( the data
for each source are independent).
The per-source marginal likelihood in the above expression is given by
\begin{align}
    p(D_n \given \bs{\alpha}) &= \int \dd \bs{w}_n \, p(D_n \given \bs{w}_n) \,
      p(\bs{w}_n \given \bs{\alpha})\\
    &= \int \dd \bs{w}_n \, p(D_n \given \bs{w}_n) \, p(\bs{w}_n \given \bs{\alpha}) \,
      \frac{p(\bs{w}_n \given D_n, \bs{\alpha}_0)}{p(\bs{w}_n \given D_n, \bs{\alpha}_0)}\\
    &= p(D_n \given \bs{\alpha}_0) \, \int \dd \bs{w}_n \,
      \frac{p(\bs{w}_n \given \bs{\alpha})}{p(\bs{w}_n \given \bs{\alpha}_0)} \,
      p(\bs{w}_n \given D_n, \bs{\alpha}_0) \label{eq:marglike} \quad .
\end{align}
Using the Monte Carlo integration approximation, \eqname~\ref{eq:marglike} can
be simplified to a sum over prior value ratios of the $K$ posterior samples in
the log-excess-variance parameter for each $n$ star
\begin{equation}
    \approx \frac{\mathcal{Z}_n}{K} \,
      \sum_k^{K} \frac{p(y_{nk} \given \bs{\alpha})}{p(y_{nk} \given \bs{\alpha}_0)}
\end{equation}
where we have canceled the other priors (over $\bs{\theta}$), and all
normalization constants appear in the constant scale factor $\mathcal{Z}_n$.

The above expresion gives the marginal likelihood of the velocity data for a single source given new hyperparameters $\bs{\alpha}$.
The full marginal likelihood is then the product of these individual likelihoods
\begin{align}
    p(\{D_n\} \given \alpha) &\propto \prod_n^N \frac{1}{K} \,
      \sum_k^{K} \frac{p(y_{nk} \given \alpha)}{p(y_{nk} \given \alpha_0)}
      \quad .
\end{align}
In practice, we evaluate the log-marginal-likelihood
\begin{align}
    \ln p(\{D_n\} \given \alpha) &\propto \sum_n^N \left[
      \ln\left( \sum_k^{K} \frac{p(y_{nk} \given \alpha)}{p(y_{nk} \given \alpha_0)} \right)
      - \ln K\right]\label{eq:lnlike}\\
    &\propto \sum_n^N \left[
      \underset{k}{\textrm{logsumexp}}\left[ \ln{p(y_{nk} \given \alpha)} - \ln{p(y_{nk} \given \alpha_0)} \right]
      - \ln K\right]
\end{align}
where $\textrm{logsumexp}$ (the log-sum-exp trick) provides a more stable
estimate of the sum in \eqname~\ref{eq:lnlike}.

>>>>

@@

@@

% APOGEE RV var co-authors
\author{Nathan De Lee}
\affiliation{Department of Physics, Geology, and Engineering Technology, Northern Kentucky Univerity, Highland Heights, KY 41099}
\affiliation{Department of Physics & Astronomy, Vanderbilt University, Nashville, TN 37235}

@@

\author{Nicholas Troup}
\affiliation{Department of Physics, Salisbury University, Salisbury, MD 21801}

@@

\begin{abstract}\noindent % trust me
% Context
Multi-epoch radial velocity measurements of stars can be used to identify
stellar, sub-stellar, and planetary-mass companions.
Even a small number of observation epochs can be informative about
companions, though there can be multiple qualitatively different orbital
solutions that fit the data.
We have custom-built a Monte Carlo sampler (\thejoker) that delivers reliable
(and often highly multi-modal) posterior samplings for companion orbital
parameters given sparse radial-velocity data.
% Aims
Here we use \thejoker\ to perform a search for companions to \nstars\ red-giant
stars observed in the \apogee\ survey (\DR) with $\geq 3$ spectroscopic epochs.
% Methods
We select stars with probable companions by making a cut on our posterior belief
about the amplitude of the stellar radial-velocity variation induced by the
orbit.
% Results
We provide (1)~a catalog of \nunimodal\ companions for which the stellar
companion properties can be confidently determined, (2)~a catalog of \nhighK\
stars that likely have companions, but would require more observations to
uniquely determine the orbital properties, and (3)~posterior samplings for the
full orbital parameters for all stars in the parent sample.
We show the characteristics of systems with confidently determined companion
properties and highlight interesting systems with candidate compact object
companions.
\end{abstract}

@@

Related to the above, we assume that all RV variations are
gravitational---\emph{Kepler}.
We see tentative evidence for surface oscillations at the upper giant branch
(see \figurename~\ref{fig:logg-teff-p-e}, lower left), which we presently treat
as excess variance or intrinsic jitter.
Further observations are needed to determine whether these ``systems'' have
incorrect stellar parameters, or indeed show surface oscillations related to
asteroseismic modes.

@@

\software{
    \package{Astropy} (\citealt{Astropy-Collaboration:2013}),
    \package{emcee} (\citealt{Foreman-Mackey:2013}),
    \package{IPython} (\citealt{Perez:2007}),
    \package{matplotlib} (\citealt{Hunter:2007}),
    \package{numpy} (\citealt{Van-der-Walt:2011}),
    \package{scikit-learn} (\citealt{Pedregosa:2011}),
    \package{scipy} (\url{https://www.scipy.org/}),
    \package{schwimmbad} (\citealt{Price-Whelan:2017a}),
    \package{sqlalchemy} (\url{https://www.sqlalchemy.org/}),
    \package{thejoker} (\citealt{Price-Whelan:2017b}).
    % \package{twobody} (\todo{twobody zenodo}).
}

@@

For each $n$ of $N$ RGB stars in APOGEE, we obtain $M$ posterior samples over
primary orbital parameters $\bs{\theta} = (P, e, \omega, M_0, K, v_0)$ and the
excess-variance parameter, $y = \ln s^2$, using \thejoker; For brevity in
expressions below, we will use the vector
\begin{equation}
    \bs{w} = (\bs{\theta}, y)
\end{equation}
to represent the full set of parameters.
To obtain this sampling, we use an interim (Gaussian) prior on the
excess-variance parameter parametrized by a mean and standard deviation, i.e.
$\alpha_0 = (\mu_{y,0}, \sigma_{y,0})$ as described above.
For a given source, the posterior samples in the above parameters are drawn from
the distribution
\begin{equation}
    \bs{w}_m \sim p(\bs{w}_m \given D, \bs{\alpha}_0)
\end{equation}
where $D$ represents the data for a given object.

We want to compute the likelihood of all data from all $N$ stars, $\{D_n\}$,
given a new set of hyperparameters $\bs{\alpha}$
\begin{equation}
    p(\{D_n\} \given \bs{\alpha}) = \prod_n^N p(D_n \given \bs{\alpha})
\end{equation}
where in the above, we have assumed that this likelihood is separable ( the data
for each source are independent).
The per-source marginal likelihood in the above expression is given by
\begin{align}
    p(D_n \given \bs{\alpha}) &= \int \dd \bs{w}_n \, p(D_n \given \bs{w}_n) \,
      p(\bs{w}_n \given \bs{\alpha})\\
    &= \int \dd \bs{w}_n \, p(D_n \given \bs{w}_n) \, p(\bs{w}_n \given \bs{\alpha}) \,
      \frac{p(\bs{w}_n \given D_n, \bs{\alpha}_0)}{p(\bs{w}_n \given D_n, \bs{\alpha}_0)}\\
    &= p(D_n \given \bs{\alpha}_0) \, \int \dd \bs{w}_n \,
      \frac{p(\bs{w}_n \given \bs{\alpha})}{p(\bs{w}_n \given \bs{\alpha}_0)} \,
      p(\bs{w}_n \given D_n, \bs{\alpha}_0) \label{eq:marglike} \quad .
\end{align}
Using the Monte Carlo integration approximation, \eqname~\ref{eq:marglike} can
be simplified to a sum over prior value ratios of the $M$ posterior samples in
the log-excess-variance parameter for each $n$ star
\begin{equation}
    \approx \frac{\mathcal{Z}_n}{M} \,
      \sum_m^{M} \frac{p(y_{nm} \given \bs{\alpha})}{p(y_{nm} \given \bs{\alpha}_0)}
\end{equation}
where we have canceled the other priors (over $\bs{\theta}$), and all
normalization constants appear in the constant scale factor $\mathcal{Z}_n$.

The above expresion gives the marginal likelihood of the velocity data for a single source given new hyperparameters $\bs{\alpha}$.
The full marginal likelihood is then the product of these individual likelihoods
\begin{align}
    p(\{D_n\} \given \alpha) &\propto \prod_n^N \frac{1}{M} \,
      \sum_m^{M} \frac{p(y_{nm} \given \alpha)}{p(y_{nm} \given \alpha_0)}
      \quad .
\end{align}
In practice, we evaluate the log-marginal-likelihood
\begin{align}
    \ln p(\{D_n\} \given \alpha) &\propto \sum_n^N \left[
      \ln\left( \sum_m^{M} \frac{p(y_{nm} \given \alpha)}{p(y_{nm} \given \alpha_0)} \right)
      - \ln M\right]\label{eq:lnlike}\\
    &\propto \sum_n^N \left[
      \underset{k}{\textrm{logsumexp}}\left[ \ln{p(y_{nm} \given \alpha)} - \ln{p(y_{nm} \given \alpha_0)} \right]
      - \ln M\right]
\end{align}
where $\textrm{logsumexp}$ (the log-sum-exp trick) provides a more stable
estimate of the sum in \eqname~\ref{eq:lnlike}.
